Synthesizing Quotient Lenses

### 2018-03-14 papers/QuotientLenses/main.tex
introduct Optician_Q as a shorthand for optician enhanced with QREs

%\newcommand{\QOpt}{Optician_Q}
\newcommand{\QOpt}{QRE-enhanced Optician}
\newcommand{\OpticianRuntime}{\textbf{Optician}}
\newcommand{\QREOptician}{\textbf{Optician\textsubscript{Q}}}
\newcommand{\SystemOnOptician}{\textbf{QO}}
\newcommand{\SystemOnBenchmarks}{\textbf{QQ}}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

\section{Quotient Regular Expressions}
\label{QRE}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section formalizes the set of QRE combinators that we introduced informally
in \cref{sec:example}.
\iflastminute 
\saz{Why does the ``cref'' macro not capitalize Section?}
\bcp{Dunno.  Because SIGPLAN does not want it to??  But maybe we should be
  using autoref instead anyway...}
\fi

\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\newcommand{\bsep}{\ \ \sep{} \ \ }
\begin{align*}
Q := \; & \normalize{R_1}{R_2}{f} \bsep{} \id(R) \bsep \project{R}{s} \bsep \squash{Q_1}{Q_2}{f} \\
 | \; & \perm{Q_1, \ldots, Q_n}{Q} \bsep Q_1 \semicolon Q_2 \bsep Q_1 \cdot Q_2 \bsep (Q_1
\sep Q_2) \bsep Q^*
\end{align*}
where $Q$ ranges over QREs, 
$R$ ranges over regular expressions, 
$f$ ranges over functions between regular languages, and 
$s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, each QRE $Q$ yields four semantic
objects:

\begin{tabular}{rl}
  $W(Q)$       & A regular expression, denoting the ``whole'' of $Q$\\
  $\eqrel{Q}$  & An equivalence relation on $\mathcal{L}(W(Q))$\\
  $K(Q)$       & A regular expression, denoting the ``kernel'' of $Q$, such that
                  $\mathcal{L}(K(Q))$ forms a \\
               & complete set of representatives for $\eqrel{Q}$ \\
  $\canonize(Q)$ & 
  A ``canonizing'' function. Given any $w \in \mathcal{L}(W(Q))$,
  $\canonize(Q)(w)$ \\ & is the unique $k$ in $\mathcal{L}(K(Q))$ such
  that $k$ is equivalent to $w$ mod $\eqrel{Q}$
\end{tabular}
  % of type $\mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$  which

\noindent
Intuitivley, $W(Q)$ is the regular expression representing the
external format, while $K(Q)$ is the regular expression representing
the internal format.  The equivalence relation $\eqrel{Q}$ groups
together elements in the language of $W(Q)$ that contain the same
essential information.  The function $\canonize(Q)$ picks the
representative element from each of the resulting equivalence classes.

The well-formedness constraints for QREs ensure that these four semantic
objects fit together to form a coherent quotient $W(Q)/\eqrel{Q}$ whose
equivalence classes are determined by \canonize(Q).

@@

The relationship among the semantic objects of a QRE can be understood in terms
of the combinator \normalize{R_1}{R_2}{f}, which expresses each of these pieces
explicitly.  Its whole language is just $R_1$, its kernel language is just
$R_2$, and its canonizer is just $f$; 
its equivalence relation $\equiv$ is
determined by the fibres of $f$, so we have
$s_1 \equiv s_2 \Leftrightarrow f(s_1) = f(s_2)$ for $s_1$ and $s_2$
in $\mathcal{L}(R_1)$.

@@

\begin{prooftree}
\AxiomC{$\mathcal{L}(R_2) \subseteq \mathcal{L}(R_1)$}
\AxiomC{$f : \mathcal{L}(R_1) \longrightarrow \mathcal{L}(R_2)$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R_1}{R_2}{f}}$}
\end{prooftree}

Semantically, the \kw{normalize} QRE is universal---each of the other
combinators $Q$ is equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some
surjective, idempotent function 
$f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$.  
However,
verifying that a canonization
function $f$ is surjective and idempotent is in general undecidable.
Consequently, a programmer wishing to use \kw{normalize} must
discharge strong proof obligations, which is cumbersome in
practice.\footnote{In our implementation we allow a programmer to
use \kw{normalize} at their own risk without checking these side
conditions as an ``escape hatch'' for the case when other QRE
combinators are insufficient.\bcp{Do we use this in any of the
programs we've written?}}

The remaining QRE combinators, which we discuss next, provide simpler,
more compositional ways of building canonizers.  Nevertheless,
the \kw{normalize} combinator provides a useful guide in the design of
these combinators because it gives a sufficient condition for the
well-formedness of any potential QREs.

% particularly with respect to the strategy for synthesizing quotient
% lenses (See Section~\ref{synth}), so we use it to guide the design of
% our QRE remaining combinators, which we discuss next.

@@

Figure~\ref{fig:wk} gives the inductive definitions of the whole and
kernel languages $W(Q)$ and $K(Q)$ for all of the QRE combinators.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad\quad}l@{\quad\quad}l}

Q & W(Q) & K(Q) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{Q_1}{Q_2}{f} & W(Q_1) \sep W(Q_2) & K(Q_2) \\
\normalize{R_1}{R_2}{f} & R_1 & R_2 \\
Q_1 \; ; \; Q_2 & W(Q_1) & K(Q_2) \\
Q_1 \cdot\; Q_2 & W(Q_1) \cdot W(Q_2) & K(Q_1) \cdot K(Q_2) \\
Q_1 \sep \; Q_2 & W(Q_1) \sep W(Q_2) & K(Q_1) \sep K(Q_2) \\
Q^* & W(Q)^* & K(Q)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{Q_1, \ldots, Q_n}{Q} ) = &
\bigcup \limits_{\sigma \in S_n} W(Q_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(Q_{\sigma(n)})\\
K( \perm{Q_1, \ldots, Q_n}{Q} ) = & K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q)
\cdot K(Q_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions for QRE combinators. 
In describing regular expressions, we use the notations $\sep$ and $\bigcup$ for
alternation, the notation $\cdot$ for concatenation, and the notation
$*$ for Kleene closure.  
We use the notation $S_n$ to denote the set of all permutations of the
numbers $1$ to $n$.
}
\label{fig:wk}
\end{figure}
The \kw{squash} and \kw{permutation} combinators have the two most
interesting definitions. If $Q = \squash{Q_1}{Q_2}{f}$, then the whole language
of $Q$ is $W(Q_1) \sep W(Q_2)$ because the \kw{squash} combinator
merges the whole language $W(Q_1)$ of $Q_1$ with the whole language $W(Q_2)$ of $Q_2$.
The function $f : \mathcal{L}(W(Q_1)) \longrightarrow \mathcal{L}(W(Q_2))$,
maps $\mathcal{L}(W(Q_1))$ into $\mathcal{L}(W(Q_2))$ using $f$ and then
canonizes $W(Q_2)$ into $K(Q_2)$ using $\canonize(Q_2)$.

For the $\perm{Q_1, \ldots, Q_n}{Q}$ combinator, the whole language is the
union of languages of the form $W(Q_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(Q_{\sigma(n)})$ for any permutation $\sigma$ in $S_n$, where
$S_n$ is the set of all permutations of the numbers $1$ to $n$. Intuitively,
the \kw{permutation} combinator allows for the string to match any
permutation of the $Q_i$'s while preserving the separator $Q$ in
between each of the $Q_i$'s. The kernel of the \kw{permutation} combinator
is the language $K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q) \cdot K(Q_n)$,
because the canonical permutation is the identity permuation, with each of the
parts of the input that match $Q_i$ and $Q$ canonized into $K(Q_i)$ and $K(Q)$
respectively.

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R)) &=& id_{\mathcal{L}(R)} \\
\canonize(\project{R}{s})(w) &=& s\\[1ex]
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\[3ex]
\canonize(\normalize{R_1}{R_2}{f}) &=& f\\
\canonize(Q_1 \; ; \; Q_2) &=& \canonize(Q_2) \circ \canonize(Q_1)\\
\canonize(Q_1 \cdot Q_2) &=& \canonize(Q_1) \cdot \canonize(Q_2)\\[1ex]
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\[3ex]
\canonize(Q^*) &=& \canonize(Q)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q)(s_{n-1}))\cdot
(\canonize(Q_n)(r_n))
\end{align*}
\caption{QRE canonizers. \ksf{Solomon: please define/explain the . and *
operators on functions.}}
\label{fig:canonizers}
\end{figure}
The \kw{permutation} combinator gives rise to the most interesting
definition:
\begin{align*}\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q)(s_{n-1}))\cdot
(\canonize(Q_n)(r_n))
\end{align*}
\noindent 
which places the strings that match $Q_i$ and $Q$ according to the
canonical permutation (i.e., the identity permutation) before applying
the canonizers $\canonize(Q_i)$ and $\canonize(Q)$, respectively.

Finally, Figure ~\ref{fig:relations} gives the inductive definition of the
equivalence relation $\eqrel{Q}$, which is the set-theoretic semantics
of a QRE $Q$ as an equivalence relation on the regular language $W(Q)$.

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_{\normalize{R_1}{R_2}{f}} \; w' &\iff&
f(w)=f(w') \\
w \; \equiv_{\id(R)} \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' && \text{for all }w, w'\\
w \; \equiv_{\squash{Q_1}{Q_2}{f}} \; w' &\iff& f(w) \equiv_{Q_2} w'
\text{ or } w \equiv_{Q_2} w' \\
w \; \equiv_{Q_1 \; ; \; Q_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(Q_2)) \text{ such that } w \; \equiv_{Q_1} \; k, \; w' \;
\equiv_{Q_1} \; k', \text{ and } k \; \equiv_{Q_2} \; k'\\
w \; \equiv_{Q_1 \cdot Q_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{Q_1}
\; {r'}_1, \; r_2 \; \equiv{Q_n} \; {r'}_2\\
w \; \equiv_{Q_1 \sep Q_2} \; w' &\iff& w \; \equiv_{Q_1} \; w'
\text{ or } \; w \; \equiv_{Q_2} \; w'\\
w \; \equiv_{Q^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{Q} \; {r'}_i
\\
w \; \equiv_{\perm{Q_1, \ldots, Q_n}{Q}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{Q} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Ambiguity and Well Formed QREs}
\label{subsec:well-formed-qres}
An issue that we address before presenting the full set of inference
rules for QREs is the issue of {\em ambiguity}, which appears in the
premises of the inference rules for the regular combinators. If $Q_1$ and $Q_2$
are QREs, then when applying the regular combinators to QREs, we require that
if a string $s$ matches any of the regular expressions $W(Q_1) \cdot W(Q_2)$,
\quad $W(Q_1) \sep W(Q_2)$, \quad $W(Q_1)^*$, \quad $K(Q_1) \cdot K(Q_2)$,
\quad $K(Q_1) \sep K(Q_2)$, \quad $K(Q_1)^*$, then $s$ matches that regular
expression in only one way.

@@

We have implemented QREs and the quotient lens synthesis algorithm described
above as an extension to the Boomerang
interpreter~\cite{boomerang,quotientlenses}. 
The synthesis algorithm produces Boomerang lens values, so Boomerang
treats synthesized lenses the same as hand-written ones.
To evaluate the effectiveness of QREs, QRE lenses, and QRE lens
synthesis, we conducted experiments to answer the following questions:
%
\begin{itemize}
  \item {\bf Ease of use.} Does synthesizing QRE lenses from QREs permit an
  easier development process than writing lenses by hand?
  Does synthesizing QRE lenses from QREs permit an easier development
  process than manually writing canonizers and then synthesizing lenses between
  their canonized forms?  
  \item {\bf Performance.} Is the synthesis algorithm/implementation fast enough to be
  used as part of a standard development process?
\end{itemize}
%
All evaluations were performed on a 2.5 GHz Intel Core i7 processor with 16 GB
of 1600 MHz DDR3 running macOS High Sierra.

@@

To evaluate our QRE implementation, we adapted 39 lens synthesis tasks
from the benchmark suite of the original Optician system. These
benchmarks are a combination of custom benchmarks, benchmarks derived
from FlashFill~\cite{flashfill}, and benchmarks derived from
Augeas~\cite{augeas2}.  In the pre-QRE version of Optician, 10 of
these data formats had to be modified to work with the bijectivity
constraints that Optician required. For instance, when one
representation permits whitespace where the other does not, we
modified the original version of the benchmark to allow more whitespace,
thereby restoring bijectivity (but altering the data format). With the
new QRE support, we were able to remove these alterations.  This
experience alone suggests that QREs make the lens development process
more flexible.

@@

\begin{figure}[t]
\includegraphics{generated-graphs/asts.eps}
\caption{AST node measurements for each of the three approaches
on each of the 10 non-bijective benchmark problems.
Benchmarks are sorted in order of increasing complexity as measured by
the number of AST nodes in the source and target format descriptions. 
QRE Synthesis requires far fewer AST nodes than the other
two approaches.}
%%\caption{Count of benchmark programs definable using a given AST count. We find
%%that it takes far fewer AST nodes to define benchmark lenses using QRE
%%synthesis than with Optician without QRE synthesis or without synthesis.}
\label{fig:asts}
\end{figure}

@@

We are interested in comparing three different approaches, which vary
in the amount of synthesis used. 
In the first approach, which we call \QRESize{} for QRE Synthesis, the programmer uses
QRE lens synthesis.  She must write QRE specifications of the source and
target formats and she may give examples.
In the second approach, which we call \canonizeAndSpecSize{} for
Bijective Synthesis, the
programmer uses bijective lens synthesis \`a la the original version
of Optician.  She must write canonizers by hand, along with 
regular expressions to describe the external
representations of the source and target formats. (The internal
formats can be inferred from the canonizers.) She may also
provide examples to help in the synthesis of the bijective lens.
In the third approach, which we call \LensAndSpecSize{} for No Synthesis, the
programmer writes the lens between the source and target formats
entirely by hand, including the descriptions of the source and target
formats.

@@

Figure~\ref{fig:asts} shows each of these measures for the 10
non-bijective problems in the benchmark suite.  On
average, \canonizeAndSpecSize{} used~44\% more AST nodes
than \QRESize{}, requiring an average of~237 more AST nodes. On 
average, \LensAndSpecSize{} used~193\% more AST nodes than \QRESize{}, requiring an
average of~1267 more AST nodes. These figures suggest that introducing QREs saves
programmers significant effort compared to both pre-QRE Optician and basic
Boomerang.

>>>>

%\newcommand{\QOpt}{Optician_Q}
\newcommand{\QOpt}{Optician\textsubscript{Q}}
\newcommand{\OpticianRuntime}{\textbf{Optician}}
\newcommand{\QREOptician}{\textbf{Optician\textsubscript{Q}}}
\newcommand{\SystemOnOptician}{\textbf{QO}}
\newcommand{\SystemOnBenchmarks}{\textbf{QQ}}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

\section{Quotient Regular Expressions}
\label{QRE}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section describes in further detail the set of QRE combinators that we described
in \cref{sec:example}.
\iflastminute 
\saz{Why does the ``cref'' macro not capitalize Section?}
\bcp{Dunno.  Because SIGPLAN does not want it to??  But maybe we should be
  using autoref instead anyway...}
\fi

\subsection{Syntax of Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\newcommand{\bsep}{\ \ \sep{} \ \ }
\begin{align*}
Q := \; & \normalize{R}{R'}{f} \bsep{} \id(R) \bsep \project{R}{s} \bsep \squash{Q_1}{Q_2}{f} \\
 | \; & \perm{Q_1, \ldots, Q_n}{Q} \bsep Q_1 \semicolon Q_2 \bsep Q_1 \cdot Q_2 \bsep (Q_1
\sep Q_2) \bsep Q^*,
\end{align*}
where $Q$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, each QRE $Q$ enables us to express four semantic objects

\begin{tabular}{rl}
  $W(Q)$  & a regular expression (the ``whole'' of $Q$),\\
  $\eqrel{Q}$  & an equivalence relation on $\mathcal{L}(W(Q))$,\\
  $K(Q)$ & a regular expression (the ``kernel'' of $Q$) such that
           $\mathcal{L}(K(Q))$ forms a complete \\
             & set of representatives for $\eqrel{Q}$, and \\
  $\canonize(Q)$ & 
  a ``canonizing'' function that, given any $w \in \mathcal{L}(W(Q))$,
  yields $\canonize(Q)(w)$ \\ & as the unique $k$ in $\mathcal{L}(K(Q))$ such
  that $k$ is equivalent to $w$ mod $\eqrel{Q}$
\end{tabular}
  % of type $\mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$  which

The well-formedness constraints for QREs ensure that these four semantic
components fit together to form a coherent quotient $W(Q)/\eqrel{Q}$ whose
equivalence classes are determined by \canonize(Q).

@@

The relationship among the semantic objects of a QRE can be understood in terms
of the \normalize{R}{S}{f} combinator, which exposes each of these pieces
explicitly.  Its whole language is just $R$, its kernel language is just
$S$, and $\canonize(\normalize{R}{S}{f}) = f$.  Its equivalence relation is
determined by the fibres of $f$, so we have
$s_1 \equiv s_2 \Leftrightarrow f(s_1) = f(s_2)$.

@@

\begin{prooftree}
\AxiomC{$\mathcal{L}(S) \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(S)$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}

Semantically, the \kw{normalize} QRE is universal---each of the other
combinators $Q$ is equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some
surjective, idempotent function
$f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$.   However,
although \kw{normalize} is powerful, verifying that a canonization function $f$
is surjective and idempotent is in general undecidable.  This means that a
programmer wishing to use \kw{normalize} must discharge strong proof
obligations, which is cumbersome in practice.\footnote{In our implementation we allow a programmer to use \kw{normalize} at their
own risk without checking these side conditions as an ``escape hatch'' for the
case when other QRE combinators are insufficient.\bcp{Do we use this in any
  of the programs we've written?}}

The remaining QRE combinators provide simpler, more compositional ways of
building canonizers.  Nevertheless, the \kw{normalize} combinator gives a
sufficient condition for potential QREs to be well-behaved, particularly with
respect to the strategy for synthesizing quotient lenses which we shall
introduce in Section~\ref{synth}, so we use it to guide the design of our QRE
combinators, as we see next.

@@

Figure~\ref{fig:wk} gives the inductive definitions of the whole and
kernel languages $W(Q)$ and $K(Q)$ for all of the QRE forms.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

Q & W(Q) & K(Q) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{Q}{Q_1}{f} & W(Q) \sep W(Q_1) & K(Q_1) \\
\normalize{R}{R'}{f} & R & R' \\
Q_1 \; ; \; Q_2 & W(Q_1) & K(Q_2) \\
Q_1 \cdot Q_2 & W(Q_1) \cdot W(Q_2) & K(Q_1) \cdot K(Q_2) \\
Q_1 \sep Q_2 & W(Q_1) \sep W(Q_2) & K(Q_1) \sep K(Q_2) \\
Q^* & W(Q)^* & K(Q)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{Q_1, \ldots, Q_n}{Q} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})\\
K( \perm{Q_1, \ldots, Q_n}{Q} ) = & K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q)
\cdot K(Q_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions}
\label{fig:wk}
\end{figure}
The \kw{squash} and \kw{permutation} combinators have the two most
interesting definitions. If $Q = \squash{Q_1}{Q_2}{f}$, then the whole language
of $Q$ is $W(Q_1) \sep W(Q_2)$. This is because the \kw{squash} combinator
takes the whole language $W(Q_1)$ of $Q_1$, the whole language $W(Q_2)$ of $Q_2$
and a function $f : \mathcal{L}(W(Q_1)) \longrightarrow \mathcal{L}(W(Q_2))$,
maps $\mathcal{L}(W(Q_1))$ into $\mathcal{L}(W(Q_2))$ using $f$ and then
canonizes $W(Q_2)$ into $K(Q_2)$ using $\canonize(Q_2)$.

For the $\perm{Q_1, \ldots, Q_n}{Q}$ combinator, the whole language is the
union of languages of the form $W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})$ for any permutation $\sigma$ in $S_n$, where
$S_n$ is the set of all permutations of the numbers $1$ to $n$. This is because
the \kw{permutation} combinator allows for the string to match any
permutation the $Q_i$'s while also taking into account the separator $Q$ in
between each of the $Q_i$'s. The kernel of the \kw{permutation} combinator
is the language $K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q) \cdot K(Q_n)$,
because the canonical permutation is the identity permuation, with each of the
parts of the string that match $Q_i$ and $Q$ canonized into $K(Q_i)$ and $K(Q)$
respectively.

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R)) &=& id_{\mathcal{L}(R)} \\
\canonize(\project{R}{s})(w) &=& s\\
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\
\canonize(\normalize{R}{R'}{f}) &=& f\\
\canonize(Q_1 \; ; \; Q_2) &=& \canonize(Q_2) \circ \canonize(Q_1)\\
\canonize(Q_1 \cdot Q_2) &=& \canonize(Q_1) \cdot \canonize(Q_2)\\
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\
\canonize(Q^*) &=& \canonize(Q)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q_n)(r_n))
\end{align*}
\caption{QRE canonizers}
\label{fig:canonizers}
\end{figure}
Again, the \kw{permutation} combinator gives the most interesting
definition,
\begin{align*}\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q_n)(r_n))
\end{align*}
\noindent which, as we have explained, is defined so that the strings that match
$Q_i, Q$ are placed in the canonical permutation which is the identity
permutation before $\canonize(Q_i)$ and $\canonize(Q)$ are applied
respectively.

Finally, Figure ~\ref{fig:relations} gives the inductive definition of the
equivalence relation $\eqrel{Q}$. This is the set-theoretic semantics of QREs
as an equivalence relation on a regular language.

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
f(w)=f(w') \\
w \; \equiv_{\id(R)} \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' && \text{for all }w, w'\\
w \; \equiv_{\squash{Q_1}{Q_2}{f}} \; w' &\iff& f(w) \equiv_{Q_2} w'
\text{ or } w \equiv_{Q_2} w' \\
w \; \equiv_{Q_1 \; ; \; Q_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(Q_2)) \text{ such that } w \; \equiv_{Q_1} \; k, \; w' \;
\equiv_{Q_1} \; k', \text{ and } k \; \equiv_{Q_2} \; k'\\
w \; \equiv_{Q_1 \cdot Q_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{Q_1}
\; {r'}_1, \; r_2 \; \equiv{Q_n} \; {r'}_2\\
w \; \equiv_{Q_1 \sep Q_2} \; w' &\iff& w \; \equiv_{Q_1} \; w'
\text{ or } \; w \; \equiv_{Q_2} \; w'\\
w \; \equiv_{Q^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{Q} \; {r'}_i
\\
w \; \equiv_{\perm{Q_1, \ldots, Q_n}{Q}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{Q} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Ambiguity and Well Formed QREs}
\label{subsec:well-formed-qres}
An issue that we address before presenting the full set of inference
rules for QREs is the issue of {\em ambiguity}, which appears in the
premises of the inference rules for the regular combinators. If $Q_1$ and $Q_2$
are QREs, then when applying the regular combinators to QREs, we require that
if a string $s$ matches any of the regular expressions $W(Q_1) \cdot W(Q_2)$,
\quad $W(Q_1) \sep W(Q_2)$, \quad $W(Q_1)^*$, \quad $K(Q_1) \cdot K(Q_2)$,
\quad $K(Q_1) \sep K(Q_2)$, \quad $K(Q_1)^*$, then $s$ matches that regular
expression in only one way.

@@

We have implemented QREs and the quotient lens synthesis algorithm described
above as an extension to the Boomerang
interpreter~\cite{boomerang,quotientlenses}. We have extended the existing
Optician tool to synthesize QRE lenses from QREs. We will use word
\emph{\QOpt} to denote our extended version of Optician, and Optician to denote
the pre-QRE version of Optician. The synthesis algorithm produces Boomerang lens
values, so Boomerang treats synthesized lenses the same as hand-written ones. To
evaluate the effectiveness of QREs, QRE lenses, and QRE lens synthesis, we
conducted experiments to answer the following questions:
%
\begin{itemize}
  \item {\bf Ease of use.} Does synthesizing QRE lenses from QREs permit an
  easier development process than writing lenses by hand?
  Does synthesizing QRE lenses from QREs permit an easier development
  process than manually writing canonizers and then synthesizing lenses between
  their canonized forms?  
  \item {\bf Performance.} Is the synthesis algorithm/implementation fast enough to be
  used as part of a standard development process?
\end{itemize}
%
All evaluations were performed on a 2.5 GHz Intel Core i7 processor with 16 GB
of 1600 MHz DDR3 running macOS High Sierra.

@@

To evaluate our QRE implementation, we adapted 39 lens synthesis tasks
from the benchmark suite of the original Optician system. These
benchmarks are a combination of custom benchmarks, benchmarks derived
from FlashFill~\cite{flashfill}, and benchmarks derived from
Augeas~\cite{augeas2}.  In these 39 benchmarks, 10 of
the data formats had to be modified to work with the bijectivity
constraints that Optician required. For instance, when one
representation permits whitespace where the other does not, we
modified the original version of the benchmark to allow more whitespace,
thereby restoring bijectivity (but altering the data format). With the
new QRE support, we were able to remove these alterations.  This
experience alone suggests that QREs make the lens development process
more flexible.

@@

\begin{figure}[t]
\includegraphics{generated-graphs/asts.eps}
\caption{AST node measurements for each of the three approaches
on each of the 10 non-bijective benchmark problems.
QRE Synthesis requires far fewer AST nodes than the other
two approaches.}
%%\caption{Count of benchmark programs definable using a given AST count. We find
%%that it takes far fewer AST nodes to define benchmark lenses using QRE
%%synthesis than with Optician without QRE synthesis or without synthesis.}
\label{fig:asts}
\end{figure}

@@

We are interested in comparing three different approaches, which vary
in the amount of synthesis used. 
In the first approach, which we call \QRESize{} for QRE Synthesis, the programmer uses
QRE lens synthesis.  She must write QRE specifications of the source and
target formats and she may give examples.
In the second approach, which we call \canonizeAndSpecSize{} for
Bijective Synthesis, the
programmer uses bijective lens synthesis \`a la Optician.
She must write canonizers by hand, along with 
regular expressions to describe the external
representations of the source and target formats. (The internal
formats can be inferred from the canonizers.) She may also
provide examples to help in the synthesis of the bijective lens.
In the third approach, which we call \LensAndSpecSize{} for No Synthesis, the
programmer writes the lens between the source and target formats
entirely by hand, including the descriptions of the source and target
formats.

@@

Figure~\ref{fig:asts} shows each of these measures for the 10
non-bijective problems in the benchmark suite.  On
average, \canonizeAndSpecSize{} used~44\% more AST nodes
than \QRESize{}, requiring an average of~237 more AST nodes. On 
average, \LensAndSpecSize{} used~193\% more AST nodes than \QRESize{}, requiring an
average of~1267 more AST nodes. These figures suggest that introducing QREs saves
programmers significant effort compared to both Optician and basic
Boomerang.
